---
title: Comment on Capdevila et al. (2021) "Global patterns of resilience decline in vertebrate populations"
author:
  - Mark D. Scheuerell:
      email: scheuerl@uw.edu
      institute: wacfwru
      correspondence: true
institute:
  - wacfwru: U.S. Geological Survey Washington Cooperative Fish and Wildlife Research Unit, School of Aquatic and Fishery Sciences, University of Washington, Seattle, WA
output:
  pdf_document:
    highlight: haddock
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
documentclass: article
geometry: margin=1in
linestretch: 1.1
fontsize: 12pt
header-includes:
  - \usepackage{lineno}
  - \linenumbers
bibliography: "references.bib"
csl: "ecology-letters.csl"
---

```{r set_options, echo = FALSE, message = FALSE}
options(width = 100)
knitr::opts_chunk$set(message = FALSE, warning = FALSE)

## load the package
library("knitcitations")

## clean out any existing .bib files
cleanbib()

## set pandoc for formatting
options("citation_format" = "pandoc")
```

\vspace{0.2in}

This is version `r paste0('0.',format(Sys.time(), '%y.%m.%d'))`.

# Background

`r citet("10.1111/ele.13927")` use data from the Living Planet Database `r citep("10.1098/rstb.2004.1584")`, to evaluate temporal trends in what they refer to as resistance and recovery. Given a time series of counts, Capdevila et al. (2021) first calculate the change in population size from one time step to another as

\begin{equation}
r_t = \log \left( \frac{N_{t+1}}{N_t} \right),
\end{equation}

such that $r$ indicates whether a population is increasing $(r > 0)$, decreasing $(r < 0)$ or stable $(r = 0)$. They then refer to times when $r > 0$ as instances of recovery, and those when $r < 0$ as periods of resistance. Once the estimates of $r$ have been calculated, Capdevila et al. create two different time series of recovery and resistance.

To estimate possible trends in recovery and resistance over time, Capdevila et al. use a state-space model for a biased random walk observed with error, whereby the true underlying change in population size is given by

\begin{equation}
r_t = r_{t-1} + \mu + E_t,
\end{equation}

where $\mu$ is the upward or downward bias over time and $E_t \sim \text{N}(0, \sigma^2)$. The observed (estimated) change in population size $(Y_t)$ is then given by

\begin{equation}
Y_t = r_t + F_t,
\end{equation}

where $F_t \sim \text{N}(0, \tau^2)$\footnote{I note here that Capdevila et al. incorrectly refer to $Y_t$ as the true change in population size, rather than $r_t$.}. Capdevila then claim that they can rearrange equation (3) and substitute it into equation (2), as apparently `r citet("10.1126/science.aba1289")` did, to arrive at

\begin{equation}
Y_t = Y_{t-1} + \mu + E_t + F_t.
\end{equation}

However, their algebra is incorrect and the combined equations should instead be

\begin{equation}
Y_t = Y_{t-1} + \mu + E_t + F_t - F_{t-1}.
\end{equation}

Here I show that this approach fails to recover the true changes in the log of population size from time step to another, and offer a different, but related, approach to do so.


# Estimating population size

There is a long history of estimating changes in population size from time series of count data, much of which is based upon a stochastic, discrete-time Gompertz model `r citep(list("10.2307/2937041", "10.1073/pnas.081055898"))`. Beginning with the deterministic version of the model, 

\begin{equation}
n_t = n_{t-1} \exp[a + (b - 1) \log (n_{t-1}) ],
\end{equation}

where $n_t$ is the population size at time $t$, $a$ is the intrinsic rate of population growth, and $b$ is the strength of density dependence. When $b = 1$ the change in population size is density independent, and the strength of density dependence increased as $b \rightarrow 0$. On a log scale where $x_t = \log(n_t)$,

\begin{equation}
\begin{aligned}
x_t &= x_{t-1} + a + (b - 1) x_{t-1} \\
    &= a + b x_{t-1}.
\end{aligned}
\end{equation}

Here the equilibrium population size is given by $a / (1 - b)$ for $b \neq 1$. The stochastic version of this model is a simple, first-order autoregressive process, AR(1), given by

\begin{equation} \label{proc}
x_t = a + b x_{t-1} + e_t,
\end{equation}

and $e_t \sim \text{N}(0, \sigma^2)$.

It is well known that sampling or observation errors will lead to an estimate of $b$ that is biased low, suggesting the population is under greater density dependence than it is in reality. Thus, equation (\ref{proc}) is often combined with an explicit observation model to form a state-space model, whereby the observed log-counts $(y_t)$ are a function of the true population size plus some error, such that

\begin{equation} \label{obs}
y_t = x_t + v_t,
\end{equation}

and $v_t \sim \text{N}(0, \tau^2)$. Thus, rather than estimate the changes in population size from one time step to another from the observed counts, as Capdevila did with equation (1), the correct way to do so would be to fit the state-space model given by equations (\ref{proc}) and (\ref{obs}), and then use the estimates of $x_t$ to examine changes population size over time, where

\begin{equation}
r^*_t = \left( \frac{x_{t+1}}{x_t} \right).
\end{equation}


# Other issues to consider

* length of time series (Capdevila claim 5 years is enough)

* missing data


# Simulation study

## Data

Here I simulate some data to demonstrate the shortcoming of the method of Capdevila et al.

```{r sim_data}
## set random seed for reproducibility
set.seed(666)
## number of years of data
tt <- 20
## strength of density dependence
bb <- 0.7
## instrinsic growth rate
aa <- 0.2
## SD of process errors
qq <- sqrt(0.3)
## SD of observation errors
rr <- sqrt(0.6)
## time series of process errors
ee <- rnorm(tt, 0, qq)
## create time series of log-counts
xx <- rep(NA, tt)
xx[1] <- aa
for(t in 2:tt) {
  xx[t] <- aa + bb * xx[t-1] + ee[t]
}
## observed data
yy <- rnorm(tt, xx, rr)
```

\vspace{0.25in}

```{r plot_data, echo = FALSE, fig.dim = c(6, 4), fig.cap = "Time series of simulated true log-counts (gray) and the observed values (blue)."}
## plot of states and observations
plot.ts(xx, ylim = range(xx, yy), lwd = 2, las = 1,
        col = "darkgray",
        ylab = expression(paste(italic(x[t]), "  or  ", italic(y[t]))))
lines(yy, lwd = 2, col = "blue")
```

## Model fitting

I first fit the model described by Capdevila et al. and then fit the model described above and given by equations (\ref{proc}) and (\ref{obs}).

```{r fit_model, message = FALSE, warning = FALSE}
## time series of Capdevila estimates of r
rC <- yy[-1] - yy[-tt]
## time series of estimated recovery and resistance
rec_C <- res_C <- rC
rec_C[rC < 0] <- NA
res_C[rC > 0] <- NA
  
## time series of true r
rT <- xx[-1] - xx[-tt]
## time series of true recovery and resistance
rec_T <- res_T <- rT
rec_T[rT < 0] <- NA
res_T[rT > 0] <- NA

## correlation between Capdevila estimates and truth
round(cor(rec_C, rec_T, use = "pairwise.complete.obs"), 2)
round(cor(res_C, res_T, use = "pairwise.complete.obs"), 2)


## load library
library("MARSS")

## define model structure
mod_list <- list(
  B = matrix("b"),
  U = matrix("u"),
  Q = matrix("q"),
  Z = matrix(1),
  A = matrix(0),
  R = matrix("r")
)

## define control params
con_list <- list(maxit = 2000)

mod_fit <- MARSS(matrix(yy, nrow = 1),
                 model = mod_list,
                 control = con_list)

xfit <- t(mod_fit$states)
## time series of proposed estimates of r
rP <- xfit[-1] - xfit[-tt]
## time series of estimated recovery and resistance
rec_P <- res_P <- rP
rec_P[rP < 0] <- NA
res_P[rP > 0] <- NA
  
## correlation between Capdevila estimates and truth
round(cor(rec_P, rec_T, use = "pairwise.complete.obs"), 2)
round(cor(res_P, res_T, use = "pairwise.complete.obs"), 2)


## fit Capdevila models
## recovery
# recov_fit <- MARSS(matrix(rec_C, nrow = 1),
#                    model = mod_list,
#                    control = con_list)
## resistance
# resist_fit <- MARSS(matrix(res_C, nrow = 1),
#                     model = mod_list,
#                     control = con_list)

```


# References

```{r, echo = FALSE, message = FALSE}
## create ref cited section
write.bibtex(file = "references.bib")
```

